A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-01-18 doc/large_strain.tex
mention that breakdown is done on smaller meshes
remove a couple of todo

As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
{\color{red}
In those graphs we report the memory needed to cache additional quantities required for the tangent operator here considered,
as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
from unit to real cell \cite{kronbichler12}.
Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{rerun RRZE and IWR}\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}

@@

When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
we observe a marginal increase in computational intensity and almost unchanged performance.
This coincides with our expectation that the later algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
On another hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} \todo{rerun and check}),
however the ``scalar'' algorithm performs a large number of repetitive operations that are
independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.

@@

In order to have a better understanding of the behavior of the most promising ``tensor4'' implementation,
we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
(i) `read / write' denotes reading and writing from/to a global vector (step 1 and 13);
(ii) `sum factorization' denotes application of sum factorization techniques (steps 4, 5 and 12);
(iii) `quadrature loop' denotes operations performed on each quadrature (steps 6 -- 11).
It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that
`read / write' steps takes negligible amount of time spent on each element.
For higher order bases the majority of time is spent within the quadrature loop.
When analyzing the Roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
In this case the quadrature loop consists of, essentially, a
single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
Further optimization of \texttt{Tensor} classes in \textit{deal.II} are required to
improve the performance.
Such studies, however, are beyond the scope of the current work.
}

>>>>

As outlined in the introduction, it is not only the performance of the matrix-free vector multiplication, but also the memory requirement to store a sparse matrix which is the driving force behind matrix-free methods.
Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3} demonstrate that even with caching one fourth-order symmetric tensor and one second-order tensor at each quadrature point, the matrix-free approach takes much less memory than its matrix-based counterpart.
{\color{red}
In those graphs we report the memory needed to cache additional quantities required for the tangent operator here considered,
as well as the memory required by the \texttt{deal.II} \textit{MatrixFree} object, which caches the inverse of the Jacobian of the transformation
from unit to real cell \cite{kronbichler12}.
Note that for Algorithm \ref{alg:mf_scalar} we need to evaluate gradients with respect to both deform and undeformed
configurations, and therefore two inverses will be cached via two different \textit{MatrixFree} objects.\todo{rerun IWR}\todo{this is roughly main memory transfer, do we elaborate on this/use it?}}

@@

When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
we observe a marginal increase in computational intensity and almost unchanged performance.
This coincides with our expectation that the later algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
On another hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
however the ``scalar'' algorithm performs a large number of repetitive operations that are
independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.

@@

In order to have a better understanding of the behavior of the most promising ``tensor4'' implementation,
we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
(i) `read / write' denotes reading and writing from/to a global vector (step 1 and 13);
(ii) `sum factorization' denotes application of sum factorization techniques (steps 4, 5 and 12);
(iii) `quadrature loop' denotes operations performed on each quadrature (steps 6 -- 11).
Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
global refinement in 2D and no global mesh refinement in 3D.
It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that
`read / write' steps takes negligible amount of time spent on each element.
For higher order bases the majority of time is spent within the quadrature loop.
When analyzing the Roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
In this case the quadrature loop consists of, essentially, a
single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
Further optimization of \texttt{Tensor} classes in \textit{deal.II} are required to
improve the performance.
Such studies, however, are beyond the scope of the current work.
}
